---
title: "Session5_PCA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Principal Component Analysis (PCA)
## what is PCA???
## prcomp()
#we will draw and generate PCA plots, determine how much variation each principal component accounts for 
# examine loadings (or loading scores) to determine what variables have the largest effect on the graph
## ok lets's go
## You can also download this file from the class website!

```{r}
mydata <- read.csv("https://tinyurl.com/expression-CSV", row.names=1) #samples are columns, rows are genes
### using nrow(mydata) and ncol(mydata) you can see there are 10 cols and 100 rows
#time to implement PCA
#prcomp() expectssamples to be rows and genes to be columns (why????), so we gotta transpose the matrix with t()
mydata_t<-t(mydata)
#nice!
pca<-prcomp(mydata_t, scale = TRUE)
### check it out
attributes(pca)
###
```
```{r}
### let's look closer
pca$x[,1]
pca$x[,2]
#plot these??
plot(pca$x[,1], pca$x[,2]) ### see the larger variance with 1 compared to 2
```
#
```{r}
#ok we gotta look at out PCAs
summary(pca) #notice how PC1 has 92% of the proportion of variance
```

#
```{r}
## lets do PCA
pca <- prcomp(mydata_t, scale=TRUE)
## A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2])
## Variance captured per PC
pca.var <- pca$sdev^2 
pca.var.per <- round(pca.var/sum(pca.var)*100, 1) 
pca.var.per
plot(pca)
```
```{r}
#let's colors the KO vs WT
plot(pca$x[,1:2], col=c("red","red","red","red","red", "blue","blue","blue","blue","blue")) #but this assumes i know that all the WT data comes before the KO so i can just list the color for each point
### that's really inefficient and ridiculous
```


```{r}
wt_ko<- substr(colnames(mydata),1,2) #idk what this does???
wt_ko
###you can make the plot beter with this i guess?
colvec<- as.factor(wt_ko) ### idk how this specified color
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 

```
#MY TURN AGAIN load the FILE
```{r}
x<-read.csv("UK_foods.csv")
summary(x)
dim(x)
```

#preview the first six rows
```{r}
head(x)
```
#let's  index it 
```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
#i dont know what we did this makes no sense
```


###now let's index it from the start bc using that -1 is ANNOYING AND CAUSES PROBLEMS
```{r}
x <- read.csv("UK_foods.csv", row.names=1)
head(x)
#i dont get this either
```
#barblot it
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x))) #idk how this happened
```
#try a different plot

### and another way
```{r}
pairs(x, col=rainbow(10), pch=16)
```

